---
title: "Data Analysis Project"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(janitor) 
library(tidyverse)
library(tidycensus)
library(tigris)
```

# Data Analysis Project

## Maryland 2024 Election Data

```{r}
#loading & cleaning md 2024 data
md24_raw <- read_csv("data/md_election_results_2024.csv") |>
  clean_names()

#renaming major candidate columnns & combining all write in columns
md24_with_third_party_breakdown <- md24_raw |>
  rename(
    harris_walz24 = kamala_d_harris_and_tim_walz_democratic,
    trump_vance24 = donald_j_trump_and_jd_vance_republican,
    oliver_ter_maat24 = chase_oliver_and_mike_ter_maat_libertarian,
    stein_ware24 = jill_ellen_stein_and_rudolph_ware_green,
    kennedy_shanahan24 = robert_f_kennedy_jr_and_nicole_shanahan_unaffiliated
    ) |>
  mutate(
    write_ins24 = (sharon_e_harris_write_in_democratic + ralph_jaffe_write_in_democratic + paij_boring_and_jaimey_rutschman_write_in_republican + sethatina_newman_write_in_republican + carol_asher_write_in_other_candidates + jokie_beckett_jr_write_in_other_candidates + claudia_de_la_cruz_and_karina_garcia_write_in_other_candidates + john_r_gibb_jr_write_in_other_candidates + joseph_kishore_write_in_other_candidates + william_p_stodden_write_in_other_candidates + shiva_ayyadurai_and_crystal_ellis_write_in_unaffiliated + president_r_boddie_write_in_unaffiliated + theo_brown_sr_write_in_unaffiliated + susan_maude_buchser_write_in_unaffiliated + john_cheng_and_wayne_waligorski_write_in_unaffiliated + cherunda_fox_write_in_unaffiliated + garry_hubbard_and_elizabeth_seymour_write_in_unaffiliated + shondra_irving_write_in_unaffiliated + douglas_jenkins_write_in_unaffiliated + brian_larry_kienitz_write_in_unaffiliated + richard_mayers_write_in_unaffiliated + andre_ramon_mc_neil_sr_write_in_unaffiliated + future_madam_potus_and_jessica_kennedy_write_in_unaffiliated + benjamin_l_schwalb_write_in_unaffiliated + peter_sonski_write_in_unaffiliated + lester_h_waller_jr_write_in_unaffiliated + cornel_west_and_melina_abdullah_write_in_unaffiliated + michael_yost_write_in_unaffiliated + other_write_ins)
  )

#deselecting write-in candidate columns
md24_with_third_party_breakdown <- md24_with_third_party_breakdown |>
  select(-sharon_e_harris_write_in_democratic, -ralph_jaffe_write_in_democratic, -paij_boring_and_jaimey_rutschman_write_in_republican, -sethatina_newman_write_in_republican, -carol_asher_write_in_other_candidates, -jokie_beckett_jr_write_in_other_candidates, -claudia_de_la_cruz_and_karina_garcia_write_in_other_candidates, -john_r_gibb_jr_write_in_other_candidates, -joseph_kishore_write_in_other_candidates, -william_p_stodden_write_in_other_candidates, -shiva_ayyadurai_and_crystal_ellis_write_in_unaffiliated, -president_r_boddie_write_in_unaffiliated, -theo_brown_sr_write_in_unaffiliated, -susan_maude_buchser_write_in_unaffiliated, -john_cheng_and_wayne_waligorski_write_in_unaffiliated, -cherunda_fox_write_in_unaffiliated, -garry_hubbard_and_elizabeth_seymour_write_in_unaffiliated, -shondra_irving_write_in_unaffiliated, -douglas_jenkins_write_in_unaffiliated, -brian_larry_kienitz_write_in_unaffiliated, -richard_mayers_write_in_unaffiliated, -andre_ramon_mc_neil_sr_write_in_unaffiliated, -future_madam_potus_and_jessica_kennedy_write_in_unaffiliated, -benjamin_l_schwalb_write_in_unaffiliated, -peter_sonski_write_in_unaffiliated, -lester_h_waller_jr_write_in_unaffiliated, -cornel_west_and_melina_abdullah_write_in_unaffiliated, -michael_yost_write_in_unaffiliated, -other_write_ins)


#creating new simplified data frame with combined third-party vote
md24 <- md24_with_third_party_breakdown |>
  mutate(
    third_party24 = oliver_ter_maat24 + stein_ware24 + kennedy_shanahan24
  )

#deselecting third party candidate columns
md24 <- md24 |>
  select(-oliver_ter_maat24, -stein_ware24, -kennedy_shanahan24)
```

```{r}
#exploration

#calculating total & percentages
md24 <- md24 |>
  mutate(
    total_votes24 = harris_walz24 + trump_vance24 + write_ins24 + third_party24,
    pct_harris_walz24 = harris_walz24/total_votes24*100,
    pct_trump_vance24 = trump_vance24/total_votes24*100,
    pct_write_ins24 = write_ins24/total_votes24*100,
    pct_third_party24 = third_party24/total_votes24*100
  )

glimpse(md24)
```

```{r}
#loading & cleaning md 2020 data
md20_raw <- read_csv("data/md_election_results_2020.csv") |>
   clean_names()

#renaming major candidate columnns & combining all write in columns
md20_with_third_party_breakdown <- md20_raw |>
  rename(
    biden_harris20 = joe_biden_and_kamala_harris_democratic,
    trump_pence20 = donald_j_trump_and_michael_pence_republican,
    jorgensen_cohen20 = jo_jorgensen_and_jeremy_spike_cohen_libertarian,
    gresham_hawkins_walker20 = howie_gresham_hawkins_and_angela_walker_green,
    segal_de_graaf20 = jerome_m_segal_and_john_de_graaf_bread_and_roses
    ) |>
  mutate(
    write_ins20 = (sharon_wallace_and_karen_m_short_write_in_democratic + dennis_andrew_ball_write_in_other + barbara_bellar_write_in_other + president_boddie_write_in_other + mary_ruth_caro_simmons_and_sherrie_dow_write_in_other + brian_carroll_write_in_other + todd_cella_write_in_other + mark_charles_and_adrian_wallace_write_in_other + phil_collins_write_in_other + roque_rocky_de_la_fuente_write_in_other + ryan_ehrenreich_and_veronica_ehrenreich_write_in_other + randall_foltyniewkz_write_in_other + tom_hoefling_and_andy_prior_write_in_other + shawn_howard_write_in_other + johnson_lee_write_in_other + susan_b_lochocki_write_in_other + brock_pierce_and_karia_ballard_write_in_other + deborah_rouse_and_sheila_cannon_write_in_other + peter_w_sherrill_write_in_other + jade_simmons_and_claudeligh_j_roze_write_in_other + kasey_wells_write_in_other + kanye_west_write_in_other + gloria_la_riva_and_sunil_freeman_write_in_unaffiliated + albert_raley_write_in_unaffiliated + benjamin_schwalb_write_in_unaffiliated + edward_shlikas_write_in_unaffiliated + other_write_ins)
  )

#deselecting write-in candidate columns
md20_with_third_party_breakdown <- md20_with_third_party_breakdown |>
  select(-sharon_wallace_and_karen_m_short_write_in_democratic, -dennis_andrew_ball_write_in_other, -barbara_bellar_write_in_other, -president_boddie_write_in_other, -mary_ruth_caro_simmons_and_sherrie_dow_write_in_other, -brian_carroll_write_in_other, -todd_cella_write_in_other, -mark_charles_and_adrian_wallace_write_in_other, -phil_collins_write_in_other, -roque_rocky_de_la_fuente_write_in_other, -ryan_ehrenreich_and_veronica_ehrenreich_write_in_other, -randall_foltyniewkz_write_in_other, -tom_hoefling_and_andy_prior_write_in_other, -shawn_howard_write_in_other, -johnson_lee_write_in_other, -susan_b_lochocki_write_in_other, -brock_pierce_and_karia_ballard_write_in_other, -deborah_rouse_and_sheila_cannon_write_in_other, -peter_w_sherrill_write_in_other, -jade_simmons_and_claudeligh_j_roze_write_in_other, -kasey_wells_write_in_other, -kanye_west_write_in_other, -gloria_la_riva_and_sunil_freeman_write_in_unaffiliated, -albert_raley_write_in_unaffiliated, -benjamin_schwalb_write_in_unaffiliated, -edward_shlikas_write_in_unaffiliated, -other_write_ins)

#creating new simplified data frame with combined third-party vote
md20 <- md20_with_third_party_breakdown |>
  mutate(
    third_party20 = jorgensen_cohen20 + gresham_hawkins_walker20 + segal_de_graaf20
  )

#deselecting third party candidate columns
md20 <- md20 |>
  select(-jorgensen_cohen20, -gresham_hawkins_walker20, -segal_de_graaf20)
```

## Data Exploration

```{r}
#exploration

#calculating total & percentages
md20 <- md20 |>
  mutate(
    total_votes20 = trump_pence20 + biden_harris20 + write_ins20 + third_party20,
    pct_trump_pence20 = trump_pence20/total_votes20*100,
    pct_biden_harris20 = biden_harris20/total_votes20*100,
    pct_write_ins20 = write_ins20/total_votes20*100,
    pct_third_party20 = third_party20/total_votes20*100
  )

glimpse(md20)
```

```{r}
#deliverable 2 write up

#our core data sets are Maryland's 2020 and 2024 election data, which we loaded and cleaned above. cleaning primarily involved condensing the third-party and write-in columns to make the data easier to analyze. our basic analysis so far has included calculating the total number of votes cast in each election and the percentages of votes cast for each party, broken down by county. each data frame has 24 rows — one for each Maryland county — and 10 columns. the 10 columns include:
  #the county name
  #the number of votes cast for the democratic candidate
  #the number of votes cast for the republican candidate
  #the number of third-party votes
  #the number of write-in votes
  #the total number of votes
  #the percentage of the total vote share that democratic, republican, third party and write-in candidates received

#there aren't any obvious flaws with our current data sets, though it's important to point out that the 2024 results haven't been certified and therefore aren't yet official. the data sets do have their limitations when it comes to our questions, though, and it is clear that we will still need to load and clean a number of supplementary data sets to actually answer the questions we posed in deliverable 1. this will, at the very least, include:
  #national election data
  #2024 Senate race data
  #abortion ballot measure data
  #census data

#however, once we load and clean those data sets, we don't foresee any real issues with answering the questions we posed.
```

## Maryland Voter Registration Data

```{r}
# load 2024 MD voter registration data
md24_registered_voters <- read_csv("data/md_voters_by_county_2024.csv") |>
  clean_names() |>
  slice(-25)

# load 2020 MD voter registration data
md20_registered_voters <- read_csv("data/md_voters_by_county_2020.csv") |>
  clean_names() |>
  slice(-25)
```

## Maryland Population Data

```{r}
# load MD voting-age population data
md_population22_raw <- get_acs(geography = "county",
              variables = c(population_over_18 = "B09021_001"),
              state = "MD",
              year = 2022) |>
  clean_names()
  
# clean data
md_population22 <- md_population22_raw |>
  rename(population_over_18=estimate) |>
  select(-geoid, -variable, -moe)
```

————————————————————————————————————————————————————

## Q1

*Question: How many registered voters were there in Maryland in 2024? What percentage of voting-age people are registered to vote? What percentage of them voted in 2024? How did this compare to 2020? How did this compare by county? How did this compare nationally?*

### Standardizing

```{r}
# standardizing data spellings
md_population22 <- md_population22 |>
  mutate(
    name=gsub(', Maryland', '', name),
    name=gsub(' County', '', name),
    name=gsub('Baltimore', 'Baltimore County', name),
    name=gsub('Baltimore County city', 'Baltimore City', name)
    ) |>
  rename(county=name)

md24 <- md24 |>
  mutate(
    jurisdiction=gsub('Saint Mary\'s', 'St. Mary\'s', jurisdiction)
  ) |>
  rename(county=jurisdiction)

md20 <- md20 |>
  mutate(
    jurisdiction=gsub('Baltimore', 'Baltimore County', jurisdiction),
    jurisdiction=gsub('Baltimore County City', 'Baltimore City', jurisdiction)
    ) |>
  rename(county=jurisdiction)

md24_registered_voters <- md24_registered_voters |>
  mutate(
    county=gsub('Saint Mary\'s', 'St. Mary\'s', county)
  ) |>
  rename(total_registered24=total) |>
  rename(registered_dem24=dem) |>
  rename(registered_rep24=rep) |>
  rename(registered_lib24=lib) |>
  rename(registered_oth24=oth) |>
  rename(registered_una24=una) |>
  rename(registered_nlm24=nlm) |>
  rename(registered_grn24=grn)

md20_registered_voters <- md20_registered_voters |>
  mutate(
    county=gsub('Saint Mary\'s', 'St. Mary\'s', county)
  ) |>
  rename(total_registered20=total) |>
  rename(registered_dem20=dem) |>
  rename(registered_rep20=rep) |>
  rename(registered_bar20=bar) |>
  rename(registered_grn20=grn) |>
  rename(registered_lib20=lib) |>
  rename(registered_wcp20=wcp) |>
  rename(registered_oth20=oth) |>
  rename(registered_una20=una)
```

### Joining

```{r}
# join 2024 data
md24_combined <- md_population22 |>
  left_join(md24, join_by(county))

md24_combined <- md24_combined |>
  left_join(md24_registered_voters, join_by(county)) 

md24_combined_totals <- md24_combined |>
  select(-harris_walz24, -trump_vance24, -write_ins24, -third_party24, -pct_harris_walz24, -pct_trump_vance24, -pct_write_ins24, -pct_third_party24, -registered_dem24, -registered_rep24, -registered_lib24, -registered_oth24, -registered_una24, -registered_nlm24, -registered_grn24)

# join 2020 data
md20_combined <- md_population22 |>
  left_join(md20, join_by(county))

md20_combined <- md20_combined |>
  left_join(md20_registered_voters, join_by(county))

md20_combined_totals <- md20_combined |>
  select(-trump_pence20, -biden_harris20,  -write_ins20, -third_party20, -pct_trump_pence20, -pct_biden_harris20, -pct_write_ins20, -pct_third_party20, -registered_dem20, -registered_rep20, -registered_bar20, -registered_grn20, -registered_lib20, -registered_wcp20, -registered_oth20, -registered_una20)
```

### County Calculations

```{r}
# calculate pcts by county 2024
md24_combined_totals <- md24_combined_totals |>
  mutate(
    pct_registered_that_voted24=total_votes24/total_registered24*100
  )

# calculate pcts by county 2020
md20_combined_totals <- md20_combined_totals |>
  mutate(
    pct_registered_that_voted20=total_votes20/total_registered20*100
  )

# compare 2024 to 2020 by county
md20_md24_combined <- md24_combined_totals |>
  left_join(md20_combined_totals, join_by(county))

md20_md24_combined <- md20_md24_combined |>
  select(-population_over_18.x, -population_over_18.y) |>
  mutate(
    diff_votes=total_votes24-total_votes20,
    diff_registered=total_registered24-total_registered20,
    diff_pct_voted=pct_registered_that_voted24-pct_registered_that_voted20
  )
```

### Statewide Calculations

```{r}
# calculate statewide figures

# calculate # of registered voters in 2024

sum(md24_combined_totals$total_registered24)

    # answer: 4,204,572

# calculate statewide % of registered voters who voted in 2024
sum(md24_combined_totals$total_votes24)/sum(md24_combined_totals$total_registered24)*100

    # answer: 70.00%

# calculate # of registered voters in 2020

sum(md20_combined_totals$total_registered20)

    # answer: 4,109,762

# calculate statewide % of registered voters who voted in 2020
sum(md20_combined_totals$total_votes20)/sum(md20_combined_totals$total_registered20)*100

    # answer: 73.90%

# statewide difference in total registered
sum(md24_combined_totals$total_registered24)-sum(md20_combined_totals$total_registered20)

    # answer: maryland had 94,810 more registered voters in 2024 compared to 2020

# statewide difference in total votes
sum(md24_combined_totals$total_votes24)-sum(md20_combined_totals$total_votes20)

    # answer: 93,742 fewer people voted in Maryland in 2024 compared to 2020
```

### Answers, Issues & Notable Findings

```{r}
# ANSWERS, ISSUES & NOTABLE FINDINGS:

    # How many registered voters were there in Maryland in 2024? 
  
        #  4,204,572
  
    # What percentage of voting-age people are registered to vote?
  
        # cannot be answered because population in some counties has grown so much since 2022 that the percentages don't even make sense. in queen anne's, for example, the calculations showed that 104% of the population was registered to vote.
  
    # What percentage of registered voters actually voted in 2024? 
  
        # 70.00%
  
    # How did this compare to 2020? 
  
        # maryland had 94,810 more registered voters in 2024 compared to 2020
  
        # however, 93,742 fewer people voted in 2024 than in 2020
  
        # the percentage of registered voters who voted in Maryland dropped 3.9 percentage points between 2020 and 2024
  
    # How did this compare by county?
  
        # the percentage of registered voters who voted increased in only five counties — Calvert, Washington, Cecil, Allegany and St. Mary's — between 2020 and 2024
  
        # PG County saw the largest drop in voter turnout among (-8.65 percentage points) between 2020 and 2024
  
        # Baltimore City had the lowest turnout among registered voters in both 2020 (59.41%) and 2024 (53.63%)
  
        # Talbot County had the highest rate of voter turnout in 2024 (80.08%)
  
        # the number of registered voters slightly decreased in Allegany County and Baltimore City between 2020 and 2024
  
    # How did this compare nationally?
  
        # cannot be answered due to lack of available data on voter registration
```

## Q2

*Question: How did Trump and Harris perform in MD? How did they perform in each county? How did they perform compared to 2020? Compared to nationally?*

### County Calculations

```{r}
# calculate per county numbers

md24_md20 <- md24 |>
  left_join(md20, join_by(county))

md24_md20_main_candidates <- md24_md20 |>
  select(-trump_vance24, -harris_walz24, -write_ins24, -third_party24, -total_votes24, -pct_write_ins24, -trump_pence20, -biden_harris20, -total_votes20, -pct_third_party24, -write_ins20, -third_party20, -pct_write_ins20, -pct_third_party20) |>
  mutate(
    diff_pct_rep = pct_trump_vance24-pct_trump_pence20,
    diff_pct_dem = pct_harris_walz24-pct_biden_harris20
  )

# create column for county winners
md24_md20_main_candidates <- md24_md20_main_candidates |>
  mutate(
    winner20 = case_when(
      pct_biden_harris20-pct_trump_pence20 >= 0.00001 ~ "DEMOCRAT",
      pct_trump_pence20-pct_biden_harris20 >= 0.00001 ~ "REPUBLICAN"),
    winner24 = case_when(
      pct_harris_walz24-pct_trump_vance24 >= 0.00001 ~ "DEMOCRAT",
      pct_trump_vance24-pct_harris_walz24 >= 0.00001 ~ "REPUBLICAN")
  )

# group by county winner 2024
md24_md20_main_candidates |>
  group_by(winner24) |>
  summarise(count=n())

    # answer: republicans won 16 counties, democrats won 8

# group by winner 2020
md24_md20_main_candidates |>
  group_by(winner20) |>
  summarise(count=n())

    # answer: republicans won 14 counties, democrats won 10
```

### Statewide Calculations

```{r}
# calculate statewide figures

# calculate pct of votes democrats won statewide in 2024
sum(md24_md20$harris_walz24)/sum(md24_md20$total_votes24)*100

    # answer: 62.18%

# calculate pct of votes republicans won statewide in 2024
sum(md24_md20$trump_vance24)/sum(md24_md20$total_votes24)*100

    # answer: 34.54%

# calculate pct of votes democrats won statewide in 2020
sum(md24_md20$biden_harris20)/sum(md24_md20$total_votes20)*100

    # answer: 65.36%

# calculate pct of votes republicans won statewide in 2020
sum(md24_md20$trump_pence20)/sum(md24_md20$total_votes20)*100

    # answer: 32.15%

# calculate difference — democrats
(sum(md24_md20$harris_walz24)/sum(md24_md20$total_votes24)*100)-(sum(md24_md20$biden_harris20)/sum(md24_md20$total_votes20)*100)

    # answer: -3.1767

# calculate difference — republicans

(sum(md24_md20$trump_vance24)/sum(md24_md20$total_votes24)*100)-sum(md24_md20$trump_pence20)/sum(md24_md20$total_votes20)*100

    # answer: 2.3897
```

### Answers, Issues & Notable Findings

```{r}
# ANSWERS, ISSUES & NOTABLE FINDINGS:

    # How did Trump and Harris perform in MD? 

        # Trump won 34.54% of the vote statewide
  
        # Harris won 62.18% of the vote statewide
  
    # How did they perform in each county? 

        # Trump won 2/3 of Maryland counties (most of the state's small, rural counties)

        # Harris won each of Maryland's seven most populous counties, and won three of them with roughly 75% of the vote or more (PG, Baltimore City & Montgomery)
  
    # How did they perform compared to 2020?

        # Democrats saw their vote share drop 3.17 percentage points statewide in 2024

        # Republicans saw their vote share rise by nearly 2.4 points

        # Democrats only gained ground in Garrett County, reddest county by over 6 points — and even then, they only increased their vote share from 21.02% in 2020 to 21.82% in 2024 

        # In one county, Charles County, both candidates won a smaller percentage of the vote than their party's 2020 candidate, though the shift slightly favored Trump

        # While this shows that Maryland meaningfully shifted to the right between 2020 and 2024, it also shows that more people shifted AWAY from Democrats than TOWARD Republicans 

        # Montgomery County saw the greatest shift away from Democrats (-4.14 points) and toward Republicans (+2.76 points), even though Democrats won the county with nearly 75% of the vote

        # Two counties, Kent and Talbot, flipped from Democrat to Republican between 2020 and 2024
  
    # How did they perform in MD compared to nationally?

        # Harris won 48.4% of the vote nationally, meaning she overperformed in MD by about 14 points
```

## Q3

*Question: How many Maryland voters voted for a third-party candidate or cast a write-in ballot? How did this compare to 2020? What were the county trends?*

### Joining

```{r}
# join md20 and md24 data sets
third_party_comparison <- md20 |>
  left_join(md24, join_by(county))

# select third party and write-in columns
third_party_comparison <- third_party_comparison |>
  select(county, write_ins20, pct_write_ins20, write_ins24, pct_write_ins24, third_party20, pct_third_party20, third_party24, pct_third_party24, total_votes20, total_votes24)
```

### County Calculations

```{r}
# county calculations
third_party_comparison <- third_party_comparison |>
mutate(
    write_in_diff=pct_write_ins24-pct_write_ins20,
    third_party_diff=pct_third_party24-pct_third_party20
  )
```

### Statewide Calculations

```{r}
# statewide calculations — third party

# caclulate statewide third-party vote in 2020
sum(third_party_comparison$third_party20)

    # answer: 55,171

# caclulate statewide third-party vote as a pct of all votes cast in 2020
sum(third_party_comparison$third_party20)/sum(third_party_comparison$total_votes20)*100

    # answer (rounded): 1.82%

# caclulate statewide third-party vote in 2024
sum(third_party_comparison$third_party24)

    # answer: 74,443

# caclulate statewide third-party vote as a pct of all votes cast in 2024
sum(third_party_comparison$third_party24)/sum(third_party_comparison$total_votes24)*100

    # answer (rounded): 2.53%

# calculate difference in number of third party votes
sum(third_party_comparison$third_party24)-sum(third_party_comparison$third_party20)

    # answer: 19,272

# calculate difference in pct of third party votes
(sum(third_party_comparison$third_party24)/sum(third_party_comparison$total_votes24)*100)-(sum(third_party_comparison$third_party20)/sum(third_party_comparison$total_votes20)*100)

    # answer (rounded): 0.71 percentage points
```

```{r}
# statewide calculations — write-ins

# caclulate statewide write-in vote in 2020
sum(third_party_comparison$write_ins20)

    # answer: 20,422

# caclulate statewide write-in vote as a pct of all votes cast in 2020
sum(third_party_comparison$write_ins20)/sum(third_party_comparison$total_votes20)*100

    # answer (rounded): 0.67%

# caclulate statewide write-in vote in 2024
sum(third_party_comparison$write_ins24)

    # answer: 21,978

# caclulate statewide write-in vote as a pct of all votes cast in 2024
sum(third_party_comparison$write_ins24)/sum(third_party_comparison$total_votes24)*100

    # answer (rounded): 0.75%

# calculate difference in number of write-in votes
sum(third_party_comparison$write_ins24)-sum(third_party_comparison$write_ins20)

    # answer: 1,556

# calculate difference in pct of write-in votes
(sum(third_party_comparison$write_ins24)/sum(third_party_comparison$total_votes24)*100)-(sum(third_party_comparison$write_ins20)/sum(third_party_comparison$total_votes20)*100)

    # answer (rounded): 0.074 percentage points
```

### Answers, Issues & Notable Findings

```{r}
# ANSWERS, ISSUES & NOTABLE FINDINGS:

    # How many Maryland voters voted for a third-party candidate or cast a write-in ballot?

        # 74,443 people (2.53% of all voters) voted for a third-party candidate

        # 21,978 people (0.75% of all voters) cast a write-in ballot

    # How did this compare to 2020?
        
        # In 2020, only 55,171 (1.82%) voted for third-party candidates, and 20,422 (0.67%) cast write-in ballots

        # This shows that there was a meaningful (0.71 percentage point) increase in third-party voting in 2024, meaning the third-party vote share increased by 40% between 2020 and 2024

        # Meanwhile, the percentage of write-in votes shifted only slightly

        # This seems to correspond with all the reporting that has pointed to general frustration with the two major candidates and the two-party system more broadly

    # What were the county trends?

        # The county-by-county breakdown shows almost unequivocally that the increase in third-party voting was sweeping. Not only did the percentage of third-party votes increase in every single county, the percentage range shifted from 1.35%-2.73% in 2020 to 1.85%-3.46% in 2024.

        # In Howard County, the vote share that third-party candidates received increased by more than double between 2020 and 2024, from 1.91% to 3.46%. The later vote share is two-thirds of a percentage point higher than the county with the next-highest third-party vote share. What's going on in Howard County that turned voters off to the major party candidates?
```

## Q4

*Question: How did the MD Senate candidates compare to their national party presidential nominees? What were the county trends?*

### Loading & Cleaning

```{r}
# load & clean md senate data
md24_senate_raw <- read_csv("data/md_senate_2024.csv") |>
  clean_names()

# rename columns & combine write ins
md24_senate <- md24_senate_raw |>
  rename(
    county = jurisdiction,
    alsobrooks = angela_alsobrooks_democratic,
    hogan = larry_hogan_republican,
    third_party_senate = mike_scott_libertarian) |>
  mutate(
    county=gsub('Saint Mary\'s', 'St. Mary\'s', county)
  ) |>
  mutate(
    write_ins_senate = patrick_j_burke_write_in_other_candidates + billy_bridges_write_in_unaffiliated + irwin_william_gibbs_write_in_unaffiliated + christy_renee_helmondollar_write_in_unaffiliated + robin_rowe_write_in_unaffiliated + other_write_ins)

# deselect individual write in columns
md24_senate <- md24_senate |>
  select(-patrick_j_burke_write_in_other_candidates, -billy_bridges_write_in_unaffiliated, -irwin_william_gibbs_write_in_unaffiliated, -christy_renee_helmondollar_write_in_unaffiliated, -robin_rowe_write_in_unaffiliated, -other_write_ins)
```

### Senate Calculations

```{r}
# calculate % by county
md24_senate <- md24_senate |>
  mutate(
    total_votes_senate = alsobrooks + hogan + third_party_senate + write_ins_senate,
    pct_alsobrooks = alsobrooks/total_votes_senate*100,
    pct_hogan = hogan/total_votes_senate*100
  )
```

### Joining

```{r}
# join senate data with presidential data
senate_comparison <- md24 |>
  left_join(md24_senate, join_by(county))

# deselect third party and write in columns for simplicity
senate_comparison <- senate_comparison |>
  select(-third_party24, -third_party_senate, -write_ins24, -write_ins_senate, -pct_third_party24, -pct_write_ins24)
```

### County Calculations: Senate vs. Presidential

```{r}
# calculate difference in number of votes per county
senate_comparison <- senate_comparison |>
  mutate(
    diff_total_votes = total_votes24-total_votes_senate)

# calculate difference in % of votes by party
senate_comparison <- senate_comparison |>
  mutate(
    diff_pct_trump_hogan = pct_trump_vance24-pct_hogan,
    diff_pct_harris_alsobrooks = pct_harris_walz24-pct_alsobrooks)

senate_comparison |>
  select(pct_harris_walz24, pct_alsobrooks, pct_trump_vance24, pct_hogan)
```

### Statewide Calculations: Senate vs. Presidential

```{r}
# calculate total votes in presidential race
sum(senate_comparison$total_votes24)

    # answer: 2,943,288

# calculate total votes in senate race
sum(senate_comparison$total_votes_senate)

    # answer: 3,021,378

# calculate difference in total votes
sum(senate_comparison$total_votes24)-sum(senate_comparison$total_votes_senate)

    # answer: -78,090

# recalculate (for ease) pct of votes per candidate — trump
sum(senate_comparison$trump_vance24)/sum(senate_comparison$total_votes24)*100

    # answer (rounded): 34.54%

# recalculate (for ease) pct of votes per candidate — harris
sum(senate_comparison$harris_walz24)/sum(senate_comparison$total_votes24)*100

    # answer (rounded): 62.18%

# calculate pct of votes per candidate — hogan
sum(senate_comparison$hogan)/sum(senate_comparison$total_votes_senate)*100

    # answer (rounded): 42.84%

# calculate pct of votes per candidate — alsobrooks
sum(senate_comparison$alsobrooks)/sum(senate_comparison$total_votes_senate)*100

    # answer (rounded): 54.64%

# calculate difference in pct of votes per candidate — trump vs. hogan
(sum(senate_comparison$trump_vance24)/sum(senate_comparison$total_votes24)*100)-(sum(senate_comparison$hogan)/sum(senate_comparison$total_votes_senate)*100)

    # answer (rounded): -8.29 percentage points

# calculate difference in pct of votes per candidate — harris vs. alsobrooks
(sum(senate_comparison$harris_walz24)/sum(senate_comparison$total_votes24)*100)-(sum(senate_comparison$alsobrooks)/sum(senate_comparison$total_votes_senate)*100)

    # answer (rounded): 7.54 percentage points

```

### Answers, Issues & Notable Findings

```{r}
# ANSWERS, ISSUES & NOTABLE FINDINGS:

    # How did the MD Senate candidates compare to their national party presidential nominees? 

        # 

    # What were the county trends?

        # 
```

## Q5

*Question: How did Maryland’s abortion ballot measure do in comparison to similar measures in other states? To the presidential candidates? How did different counties vote on the measure?*

### Loading & Cleaning

```{r}
# load all abortion ballot measure data 
abortion_measures <- read_csv("data/abortion_ballot_measures.csv") |>
  clean_names()

# load & clean maryland county-by-county abortion ballot measure data
md_question_1 <- read_csv("data/md_question_1.csv") |>
  clean_names() |>
  rename(county = jurisdiction) |>
  rename(voted_yes_q1 = "for") |>
  rename(voted_no_q1 = against) |>
  mutate(
    county=gsub('Saint Mary\'s', 'St. Mary\'s', county)
  )
```

### National Calculations

```{r}
# calculate yes/no percentages
abortion_measures <- abortion_measures |>
  mutate(
    total_votes = voted_yes + voted_no,
    pct_voted_yes = voted_yes/total_votes*100,
    pct_voted_no = voted_no/total_votes*100
  )
```

### Maryland County Calculations

```{r}
# calculate yes/no percentages by county
md_question_1 <- md_question_1 |>
  mutate(
    total_votes_q1 = voted_yes_q1 + voted_no_q1,
    pct_voted_yes_q1 = voted_yes_q1/total_votes_q1*100,
    pct_voted_no_q1 = voted_no_q1/total_votes_q1*100
  )
```

```{r}
# join data & deselect irrelevant info
md_abortion_joined <- md24 |>
  left_join(md_question_1, join_by(county)) |>
  select(-write_ins24, -third_party24, -pct_third_party24, -pct_write_ins24)

# calculate difference between total votes
md_abortion_joined <- md_abortion_joined |>
  mutate(
    diff_total_votes = total_votes24-total_votes_q1
  )

# compare pct of votes for/against q1 to pct of votes for candidates by party
md_abortion_joined <- md_abortion_joined |>
  mutate(
    diff_pct_yes_q1_harris = pct_voted_yes_q1-pct_harris_walz24,
    diff_pct_no_q1_trump = pct_voted_no_q1-pct_trump_vance24
  )
```

### Maryland Statewide Calculations

```{r}
# recalculate (for ease) total number of votes on q1
sum(md_abortion_joined$total_votes_q1)

    # answer: 2,891,538

# recalculate (for ease) total number of votes in presidential race
sum(md_abortion_joined$total_votes24)

    # answer: 2,943,288

# recalculate (for my own curiosity's sake) total number of votes in senate race
sum(senate_comparison$total_votes_senate)

    # answer: 3,021,378

# calculate difference between votes for president and votes on q1
sum(md_abortion_joined$total_votes24)-sum(md_abortion_joined$total_votes_q1)

    # answer: 51,750

# recalculate (for ease) overall pct of votes for q1
sum(md_abortion_joined$voted_yes_q1)/sum(md_abortion_joined$total_votes_q1)*100

    # answers (rounded): 76.06%

# recalculate (for ease) overall pct of votes against q1
sum(md_abortion_joined$voted_no_q1)/sum(md_abortion_joined$total_votes_q1)*100

    # answers (rounded): 23.94%

# calculate overall difference between pct of votes for q1 and pct of votes for harris
(sum(md_abortion_joined$voted_yes_q1)/sum(md_abortion_joined$total_votes_q1)*100)-(sum(md_abortion_joined$harris_walz24)/sum(md_abortion_joined$total_votes24)*100)

    # answer (rounded): 13.88 percentage points

# calculate overall difference between pct of votes against q1 and pct of votes for trump
(sum(md_abortion_joined$voted_no_q1)/sum(md_abortion_joined$total_votes_q1)*100)-(sum(md_abortion_joined$trump_vance24)/sum(md_abortion_joined$total_votes24)*100)

    # answer (rounded): -10.60 percentage points
```

### Answers, Issues & Notable Findings

```{r}
# ANSWERS, ISSUES & NOTABLE FINDINGS:

    # How did Maryland’s abortion ballot measure do in comparison to similar measures in other states? 

        # There were 11 ballot measures in 10 states that pertained to abortion, with dueling measures in Nebraska. Ten of the measures aimed to protect the right to abortion, and seven of them passed their specific state's constitutional amendment threshold (though all but one got more than 50% of the vote). 

        # Still, NOWHERE did a ballot measure perform better than in Maryland. The ballot measure received more than three-quarters of the vote. The other states with high passing margins were Nevada and New York, where  62% and 64% of voters voted to protect the right to abortion, respectively. This is monumental, because it means Maryland's ballot measure outperformed even the most successful ballot measures by more than 11 percentage points. 

        # Maryland is known to be a very liberal state, but to be fair, so is New York. So what about Maryland made this issue so salient?

    # How did it do in comparison to the presidential candidates? 

        # The data show that 51,750 fewer people voted on Question 1 than voted for president. This is interesting, because it means that the Maryland Senate race, with over 3 million ballots cast, received the most overall votes.

        # And although the Democratic Party's platform is generally seen as the most pro-reproductive freedom, the data also show that Maryland's ballot measure to protect the right to abortion WAY outperformed Harris. Looking at the vote share percentages, Question 1 passed with 76.06% of the vote, compared to Harris' 62.18% — a nearly 14-point margin. This means that a number of Republicans and third-party voters voted to protect the right to abortion.

        # This is further supported by the fact that Trump outperformed the share of "no" votes on Question 1 by 10 percentage points statewide. In other words, there had to have been many people who were willing to support Trump but were unwilling to support the party's generally anti-abortion platform. This also means, however, that many third-party and write-in voters voted "yes" on Question 1, because this does not fully account for the 14-point margin between Harris' vote share and the share of "yes" votes.

    # How did different counties vote on the measure?

        # In three counties — PG, Baltimore City and Montgomery — more people voted on Question 1 than voted for president, which is interesting.

        # Question 1 only received less than 50% of the vote in one county, Garrett County, where 75% of voters supported Trump (the data have repeatedly shown this is the state's most conservative county). Still, even there, 44% of voters voted "yes" on Question 1. And in other conservative Maryland counties like Allegany and Caroline, where upward of two-thirds of voters voted for Trump, more than 55% of voters also supported Question 1. In Cecil County, 64% of people voted for Trump, and 64% of people voted for Question 1. Assuming there were only slight changes in the voter pool, this means that even if everyone who didn't vote for Trump voted "yes," some 12,000 Trump voters would have also had to have voted yes on Question 1. This is fascinating voter behavior. It clearly didn't just happen in Maryland, though — the fact that Missouri passed a ballot measure to protect abortion shows that this is likely a larger trend. I recently read a research article that pointed out that many voters who were fearful about reproductive freedom may have been OK with voting for Trump BECAUSE abortion was on the ballot separately. Of course, our findings don't prove or disprove anything, but they certainly support the idea that this should be researched. 
```
