---
title: "lab_02"
author: "Derek Willis"
adapted by: "Daniel Trielli"
date: "2024-05-23"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense
data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

### **Q1** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most?

```{r}
climate_courses <- umd_courses |>
  filter(str_detect(title, "Climate")) |>
  filter(seats >= 1)

nrow(climate_courses)
```

```{r}
climate_courses |>
  group_by(department) |>
  summarise(
    count_climate_courses = n()
  ) |>
  arrange(desc(count_climate_courses))
```

**A1:**

a)  26 departments have offered classes with "climate" in the title

b)  The department of Atmospheric and Oceanic Science has offered the
    most climate-related classes (68)

------------------------------------------------------------------------

### **Q2** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top departments different in each result? Where is Journalism in those results?

```{r}
fall_2023_courses <- umd_courses |>
  filter(term == 202308) |>
  select(term, id, department, syllabus_count)
```

```{r}
fall_2023_courses |>
  group_by(department) |>
  summarise(
    count_fall_2023_courses = n()
  ) |>
  arrange(desc(count_fall_2023_courses))
```

```{r}
fall_2023_courses |>
  filter(syllabus_count >= 1) |>
  group_by(department) |>
  summarise(
    fall_2023_courses_with_syllabus = n(),
    sum_syllabus_count = sum(syllabus_count)
  ) |>
  arrange(desc(fall_2023_courses_with_syllabus))
```

```{r}
fall_2023_courses |>
  filter(syllabus_count >= 1) |>
  group_by(department) |>
  summarise(
    fall_2023_courses_with_syllabus = n(),
    sum_syllabus_count = sum(syllabus_count)
  ) |>
  arrange(desc(sum_syllabus_count))
```

**A2:**

a)  **Business and Management** (164), **School of Music** (137) and
    **English** (120) have the most courses in the fall 2023 term

b)  There are 125 departments with at least one syllabus posted for fall
    2023, compared to the 189 departments that have courses in the fall
    2023 term

c)  **Business and Management** has the most courses with at least one
    syllabus posted (82), but the second-most syllabuses posted overall
    (256). **Economics** - despite only having the 14th-most classes in
    fall 2023 - has the most syllabuses posted overall (331) and the
    second-most classes with at least one syllabus posted (50). The
    **School of Music**, which has the second-most classes, has the
    third-most classes with at least one syllabus posted (48) and the
    eighth-most syllabuses posted overall (92). **English**, meanwhile,
    has the third-most courses in fall 2023 but only the 43rd-most
    classes with at least one syllabus posted (8) and 36th-most
    syllabuses posted overall (23).

d)  **Journalism** has the sixth-most classes in fall 2023, but is
    ranked 38th in most classes with at least one syllabus posted and
    40th in most syllabuses posted overall (18).

------------------------------------------------------------------------

### **Q3** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time.

```{r}
maryland_expenses <- maryland_expenses |> 
  mutate(expenditure_date=mdy(expenditure_date))

head(maryland_expenses)
```

```{r}
maryland_expenses |>
  mutate(month = floor_date(expenditure_date, "month")) |>
  filter(committee_name == "Moore  Wes For Maryland") |>
  group_by(month, committee_name) |>
  summarise(
    total_amount_moore = sum(amount)) |>
  arrange(month)
```

**A3:** Wes Moore's fundraising committee generally spent between
\$210,000 and \$275,000 for the first six months of the 18-month
campaign before ramping up spending in early 2022. The committee then
spent upwards of \$500,000 a month between April 2022 and September 2022
— including over \$1 million in both June and July — before spending
\$3.1 million in the final month leading up to the election.

------------------------------------------------------------------------

### **Q4** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record indicated money spent in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).

```{r}
maryland_expenses_with_state = maryland_expenses |>
  mutate(
    spending_location = case_when(
      str_detect(address, " Maryland ") ~ "in-state",
      !str_detect(address, " Maryland ") ~ "out of state",
      .default = "no address"
    )
  )
```

```{r}
maryland_expenses_with_state |>
  group_by(spending_location) |>
  summarise(
    count_expenses = n(),
    amount = sum(amount, na.rm = TRUE)
  )
```

**A4:** Committees spent the majority of their money in-state, making
51,801 purchases worth \$77.7 million in Maryland alone. By comparison,
the committees spent \$49.5 million on out-of-state purchases. Another
\$2.1 million worth of expenses were not associated with an address.

(I used na.rm after googling how to remove NAs from my summarise
calculations)

------------------------------------------------------------------------

### **Q5** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab.

### ***How many classes focus on social justice issues? Which departments are they in? Which department offers the greatest number of courses on social justice? What level are they taught at?***

```{r}
social_justice_courses <- umd_courses |>
  filter(
    str_detect(title, "Social Justice")
  )

nrow(social_justice_courses)
```

```{r}
social_justice_courses <- social_justice_courses |>
  mutate(
    course_code_chr = gsub("[0-9]", "", id),
    course_code_num = gsub("[^0-9]", "", id)
  ) |>
  mutate(
    course_code_cat = case_when(
      course_code_num < 500 ~ "undergraduate",
      course_code_num >= 500 & course_code_num < 900 ~ "graduate"
  )
  )
```

```{r}
social_justice_courses |>
  group_by(department) |>
  summarise(
    count_sj_courses = n()
  ) |>
  arrange(desc(count_sj_courses))

social_justice_courses |>
  group_by(course_code_cat) |>
  summarise(
    count_sj_courses = n()
  )
```

**A5:**

a)  32 courses focus on social justice

b)  8 departments offer courses on social justice

c)  Geographical Sciences and Education Counseling and Personnel
    Services are tied for most social justice courses (7)

d)  20 classes are taught at the undergraduate level, 12 at the graduate
    level
