---
title: "Reverse Engineering Project"
author: "Cat Josie and Zavier"
date: "Oct 17"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this notebook, we are reverse engineering the story, [STORY NAME](STORY%20URL)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)

install.packages("readr")
library(readr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data.

```{r}
# Load required data

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

sports_political_donations <- read_csv("data/sports-political-donations.csv") |>
  clean_names()
  
sports_political_donations <- sports_political_donations |>
  mutate(
   owner = str_to_upper(owner),
   team = str_to_upper(team),
   recipient = str_to_upper(recipient),
   party = str_to_upper(party),
   amount = gsub(",", "", amount),
   amount = gsub("\\$", "", amount),
   amount = as.double(amount)
  ) 

# I have used gsub() before, but I didn't consider that "$" is an R function and used ChatGPT to figure out that the "\\" were necessary to remove them from the df using gsub().

# I gave this prompt: "hi chatgpt how do i remove $ from a dataset in R using gsub()"

# It gave this response: data$price <- gsub("\\$", "", data$price)

# I instantly realized why the $ wasn't reading before and rewrote my code with backslashes

```

## Sentences to Engineer

In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

-   **Sentence text**: “A deep search in the Federal Election Commission database of campaign finances for principal owners, controlling owners, co-owners and commissioners from the NBA, NFL, NHL, WNBA, MLB and NASCAR reveals that this deep-pocketed group has sent \$34.2 million (72.9 percent) to Republican campaigns or super PACs purely supporting Republican causes, compared to \$10.1 million (21.5 percent) to Democrats over the past three elections.”
-   **Analysis summary**: We were able to confirm the finding ... eventually. Cleaning the data to anaylze this sentence proved to be incredibly challenging due to the presence of special characters in the "amount" column data, and R's corresponding categorization of the data as character data. But once we worked through those issues, it was mostly OK. We did run into some issues creating new columns for the amount totals and percentages, but creating a new data frame with only the data from the amount column seemed to solve these issues. And, finally, we were able to use this data set to confirm that, over three election cycles, pro sports owners contributed \$34.2 million (72.9 percent) to Republicans and \$10.1 million (21.5 percent) to Democrats.

<!-- -->

-   [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

sports_political_donations_amounts <- sports_political_donations |>
  group_by(party) |>
  summarise(
    party_amounts=sum(amount)
  ) |>
  mutate(
    all_donations=sum(party_amounts)
  ) |>
  mutate(
    pct_total=(party_amounts / all_donations *100))
  ) |>
  
sports_political_donations_amounts |>
    arrange(desc(pct_total))
```

### Sentence 2

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

-   **Sentence text**: [Paste in sentence to engineer here]
-   **Analysis summary**: [Write up two to three sentences describing the results of your analysis. Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-
