---
title: "week16_recap"
author: "Daniel Trielli"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(tidytext)
library(lubridate)
library(janitor)
library(tigris)
library(tidycensus)
library(corrr)
library(rvest)
```

## Load Data

[USAFacts compiled data on inspections on child care facilities throughout the country](https://usafacts.org/data-projects/childcare-inspections) - specifically, how many of those facilities are overdue for inspection. Let's take a look on the data they have on Maryland.

First, here the data we need to load:

```{r}
childcare_inspections <- read_csv('data/childcare_inspections.csv')
childcare_inspections$fips <- as.character(childcare_inspections$fips)
```

#### **Q1** The childcare data does not have the names of the counties. Join them with the md_counties data, that can be loaded in the block below. Then inspect the dataframe you created with this join and answer this: there are two columns with county names in the new dataframe. Which one is the best to be used? (hint: it has to do with homonyms)

```{r}
# This loads the county data
md_counties <- counties() |>  
  filter(STATEFP == "24")
```

```{r}
# Join the childcare data with the county data 
inspections_with_counties <- md_counties |>
  left_join(childcare_inspections, join_by("GEOID"=="fips")) |>
  clean_names()
```

##### **A1:** There are two columns that list the county names: "name" and "namelsad." In order to avoid confusion, the column "namelsad" should always be used to identify the county name. Why? Because of Baltimore. In the "name" column, both Baltimore County and Baltimore City are listed simply as "Baltimore." This column is therefore useless, as it makes the two distinct geographies indistinguishable. The "namelsad" column, meanwhile, follows each county name with the word "county" and identifies Baltimore City as such. It is thus the preferable choice because Baltimore County and Baltimore City appear as distinct entries.

------------------------------------------------------------------------

#### **Q2** Create bar chart with the percentages of overdue childcare facility inspections per county to answer (You'll need to create a column for that calculation first). What is the county with the largest percentage of facilities with overdue inspections?

```{r}
# Calculate the percentage of overdue inspections per county

inspections_with_counties <- inspections_with_counties |>
  mutate(
    pct_overdue=overdue/total*100
  )

# Create the chart with ggplot
inspections_with_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(namelsad, pct_overdue), weight=pct_overdue)) +
  coord_flip() +
  labs(
    title="Maryland Counties by Percentage of Overdue Chilcare Inspections",
    x = "county",
    y = "percentage of overdue childcare inspections",
    caption = "source: USAFacts"
  )
```

##### **A2:** Montgomery County has the greatest percentage of childcare facilities with overdue inspection reports (65.86%).

------------------------------------------------------------------------

#### **Q3** Next, we want to see if the number of child care facilities - with overdue inspections or otherwise - is appropriate for each county. So we're going to calculate the rate of children per facility per county. To do that, you'll first need Census data to get the population of children under 5 for each county. Which county has the biggest rate of children per facility (i.e. they could use more facilities)?

```{r}
# First, grab data from the ACS. The ACS variable for children age under 5 is B06001_002 We want 2022 data.
children <- get_acs(geography = "county",
              variables = c(children_under_5 = "B06001_002"),
              state = "MD",
              year = 2022)
```

```{r}
# Rename the column name 'estimate', which has our population data, with something more useful, like 'children_pop'
children <- children |>
  rename(children_under_5=estimate) |>
  select(-variable) |>
  clean_names()
```

```{r}
# Join the dataframes
inspections_with_counties_and_population <- inspections_with_counties |>
  left_join(children, join_by(geoid)) |>
  select(-name.x, -name.y)

# Calculate the rates for children per facility
inspections_with_counties_and_population <- inspections_with_counties_and_population |>
  mutate(
    children_per_facility=children_under_5/total
  )
```

##### **A3: Garrett County has the highest child-to-childcare facility ratio with a rate of over 87 children per facility.**

------------------------------------------------------------------------

#### **Q4** Make a map of the results for question 3. Are there any areas of the state that are more lacking in childcare facilities?

```{r}
ggplot() +
  geom_sf(data=inspections_with_counties_and_population, aes(fill=children_per_facility)) +
  scale_fill_viridis_b(option="magma",trans = "log")
```

##### A4: It appears as though Western and Central Maryland are the most strapped for childcare, with most counties only having one childcare facility for every 50-70 children. This comes in stark contrast to Southern Maryland, where all but a few counties have one childcare facility for ever 30-40 children.

------------------------------------------------------------------------

#### **Q5** Now let's explore if we can find patterns in under-served areas, particularly when it comes to median income of the population. First, join the childcare data median income data from the Census. Then, create a scatterplot and perform a cor.test(), and answer: is there is any correlation between income and rate of children per facilities? What makes you conclude that?

```{r}
# Again, grab data from the ACS. The ACS variable for median income is B19013_001. We want 2022 data again.
median_income <- get_acs(geography = "county",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2022)
```

```{r}
# Rename the column name 'estimate', which has our income data, with something more useful, like 'median_income'
median_income <- median_income |>
  rename(median_income=estimate) |>
  select(-variable) |>
  clean_names()
```

```{r}
# Join the dataframes
inspections_with_counties_and_population_and_income <- inspections_with_counties_and_population |>
  left_join(median_income, join_by(geoid)) |>
  select(-name)
```

```{r}
# Create a scatterplot with median income and rate of children per facility
inspections_with_counties_and_population_and_income |>
  ggplot(aes(x = children_per_facility, y = median_income)) +
  geom_point() +
  geom_smooth(method=lm)
```

```{r}
# Create a correlation test.
cor.test(inspections_with_counties_and_population_and_income$median_income, inspections_with_counties_and_population_and_income$children_per_facility)
```

##### A5: As if the underwhelming scatterplot wasn't clear enough, the correlation test shows almost unequivocally that there is little to no correlation between median income and the number of children per childcare facility. Case in point, the r-value is a mere -0.11, meaning there is a very weak negative correlation between the variables. The t-value, meanwhile, is a very weak -0.5223 — yet another indication that the relationship between the variables is statistically meaningless. Likewise, the p-value is a sky-high 0.6067 — 12 times higher than a scientifically significant p-value of less than 0.05. In concert, these values all indicate that the correlation is both too weak and too random to be statistically relevant.

------------------------------------------------------------------------

#### **Q6** Finally, let's do some text analysis. We another dataset, with inspection reports from informal child care providers in the state. This dataset has an informative column of initial findings: a column that describes issues found in these providers, if any. Your goal is to create bigrams for this column and find common issues that are mentioned in these findings. And then answer: what kind of issues listed there you think are interesting? They don't have the most frequent.

```{r}
# Read the reports data
reports <- read_csv('data/childcare-informal-provider-inspections.csv') |> 
  clean_names()
```

```{r}
# Create a stopword dictionary (feel free to include more words)
stop_words <- stop_words |> 
  add_row(word = "with") |> 
  add_row(word = "s") |> 
  add_row(word = "h") |>
  add_row(word = "and") |>
  add_row(word = "standards") |>
  add_row(word = 'the') |> 
  add_row(word = 'a') |> 
  add_row(word = 'for') |> 
  add_row(word = 'to') |> 
  add_row(word = "not") |> 
  add_row(word = "in") |> 
  add_row(word = "non") |> 
  add_row(word = "epp") |>
  add_row(word = "items") |>
  add_row(word = "compliant")
```

```{r}
# Unnest tokens into bigrams, removing stop words and repasting bigrams, list bigrams
uniques <- reports |> 
  select(initial_findings) |>
  unnest_tokens(word, initial_findings)

view(uniques)

uniques |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

uniques |>
  count(word, sort = TRUE) |>
  top_n(25) |>
  mutate(word = reorder(word, n)) |>
  ggplot(aes(x = word, y = n)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
      labs(x = "unique words",
      y = "count",
      title = "Count of unique words found in Maryland childcare inspections")

reports_bigrams <- reports |>
  unnest_tokens(bigram, initial_findings, token = "ngrams", n = 2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(bigram = paste(word1, word2, sep=" ")) |>
  group_by(bigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

reports_bigrams
```

##### A6: The list of common bigrams shows that lack of emergency preparation is by far the most cited issue in childcare inspection reports. Case in point, six of the top 10 bigrams identified in the initial findings section relate to emergency preparation, and two of those bigrams include the word "missing." This equates to 67 individual references in total. The overwhelming prevalence of references to emergency preparation was made even clearer when I researched what "ertg" and "ecma" mean. Evidently, they refer to "emergency ready to-go" bags and "emergency care and medical authorization" documents. In other words, the bigrams "emergency bag," "emergency ready" and "ertg missing" are in reference to the same thing, as are the bigrams "ecma docs" and "emergency documents." This is incredibly interesting, because, if I were a parent, one of my chief concerns would be whether the people responsible for watching my child were prepared in the event of an emergency. I mean, if my child took medication, and the facility didn't have an emergency bag or the proper emergency medical documentation, what would happen in the event of an evacuation? That being said, the other bigrams — things like "cleaning products" and "outlet covers" — are pretty interesting too. It's kind of a no-brainer to keep cleaning products out of reach and make sure all outlets are covered, so it's pretty concerning that these are common issues at childcare facilities.
