---
title: "lab_01"
author: "Derek Willis"
adapted by: "Daniel Trielli"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

### **Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you?

```{r}
umd_courses |>
  group_by(department) |>
  summarise (
    mean_seats = mean(seats)
  ) |>
  arrange(desc(mean_seats))

```

**A1:** The courses in the Engineering, Cooperative Education department have by far the highest mean number of seats at over 300. I think it's more interesting to me that the mean number of seats for the Engineering, Cooperative Education department is *so* much higher than the department with the second-highest mean. I also find it interesting that many of the science departments — chemistry, life sciences, biological sciences, physics, etc. — have such large classes (a mean of 60+ students). With that many students in one class, how do you even learn? How can you hear the professor teach? How long does attendance take? And it's science, too — how do you do labs? It just blows my mind that anyone could learn chemistry in a room with 100 people.

------------------------------------------------------------------------

### **Q2** Find out the greatest number of times any individual class has been offered in our dataset. You must show the greatest number of classes first.

```{r}
umd_courses |>
  group_by(title) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes)) 

```

**A2:** 1,303 times — Doctoral Dissertation Research

### **Q3** Count the number of courses for each instructor and department. Which instructor has taught the most?

```{r}
umd_courses |>
  group_by(instructors, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

**A3**: Keith Yearwood

------------------------------------------------------------------------

### **Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see.

```{r}
umd_courses |>
  group_by(instructors, department) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes)) |>
  na.omit(instructors)
```

**A4:** Because there are so many courses without a listed instructor, you have to scroll through 80 rows of data before identifying Keith Yearwood as the professor who has taught the greatest number of classes. So, put simply, finding the answer to Q3 would be much easier if all the courses without a listed instructor — all the NAs — were excluded from the summarized data set.

A very quick Google search informed me about na.omit, so I tried that with the instructor column, and it seemed to work relatively well. Although the courses with "Instructor: TBA" listed as the course instructor still appear in this summarized data set, using na.omit definitely made the data considerably easier to interpret.

...

As a quick aside ... this is only true given the terms of the question, which asks for the number of courses for each instructor *and* *department*. If you include professors who teach in more than one department, Keith Yearwood has actually taught the fourth-most. Elsa Barkley Brown has actually taught the greatest number of classes in our data set (73), followed by Jason Kuo (71) and Manoj Franklin (70). However, the above calculation doesn't show this because they each teach in classes in several different departments, a few of which have been renamed and thus are categorized separately. And, technically, Manoj Franklin has taught a greater number of classes than any of them (83), but this isn't immediately clear because he has co-taught several courses.

```{r}
umd_courses |>
  group_by(instructors) |>
  summarise(
    count_classes = n()
  ) |>
  arrange(desc(count_classes))
```

------------------------------------------------------------------------

### **Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting?

**A5:**

I decided to tackle the following questions:

**"How many of the classes are 500-level classes or higher? Do graduate classes tend to be smaller, larger or around the same size as undergraduate classes? Does this differ by department?"**

The first part of this, I needed to separate the characters from the numbers in the course codes (column "id"), so I did ask ChatGPT to help.

**Here's what I put in**:

*hi chat gpt, i have a data frame in r that has both characters and numbers, how do i separate it into two columns, one with only characters and the other with only numbers?*

**Here's what it spit out**:

*To separate a column containing both characters and numbers into two separate columns (one for characters and one for numbers) in R, you can use a combination of regular expressions and the `dplyr` package (or base R if you prefer).*

*Here’s how you can do it using `dplyr` and `tidyr`:*

### *Example Data*

``` r
# Sample data frame
df <- data.frame(mixed_column = c("abc123", "def456", "ghi789"))
```

### *Using `dplyr` and `tidyr`*

``` r
library(dplyr)
library(tidyr)

# Separate characters and numbers into two columns
df_separated <- df %>%
  mutate(char_column = gsub("[0-9]", "", mixed_column),  # Remove numbers to extract characters
         num_column = as.numeric(gsub("[^0-9]", "", mixed_column)))  # Remove characters to extract numbers

print(df_separated)
```

### *Explanation:*

1.  ***`gsub("[0-9]", "", mixed_column)`**: This removes all the numbers from the `mixed_column`, leaving only the characters.*
2.  ***`gsub("[^0-9]", "", mixed_column)`**: This removes all non-numeric characters, leaving only the numbers.*
3.  ***`as.numeric()`**: This converts the extracted numeric strings into actual numbers.*

### *Output:*

``` r
  mixed_column char_column num_column
1       abc123         abc        123
2       def456         def        456
3       ghi789         ghi        789
```

*If you prefer to use **base R** instead:*

``` r
df$char_column <- gsub("[0-9]", "", df$mixed_column)
df$num_column <- as.numeric(gsub("[^0-9]", "", df$mixed_column))

print(df)
```

*This should give you the desired result of separating characters and numbers into two columns.*

From this information I learned:

-   The function gsub() function will substitute a string for another string

-   The gsub() function requires a pattern, a replacement and a vector

Using all that info, I used this code to create a new data set called "umd_separated" with new categories called "course_code_chr" and "course_code_num"

```{r}
umd_separated <- umd_courses |>
  mutate(
    course_code_chr = gsub("[0-9]", "", id),
    course_code_num = gsub("[^0-9]", "", id)
  )
```

-   The course_code_chr = gsub("[0-9], "", id) means:

    -   create a column called "course_code_chr"

    -   the pattern "[0-9]" tells R to look at all the numbers

    -   the replacement "" tells R to replace them all with nothing — essentially telling R to delete them, leaving just the characters

    -   the vector "id" tells R to do this all to the data in the "id" column

-   Likewise, course_code_num = gsub("[\^0-9]", "", id) tells R to look at all the non-number characters in the "id" column, replace them with nothing and put what remains — so, just the course code numbers — in a column called "course_code_num"

I then used the textbook to learn how to do the second part of this.

Chapter 11.2 of the textbook includes this example for the case_when() function:

```{r}
maryland_winred_categories <- maryland_winred_contributions |>
  mutate(
    amount_category = case_when(
        amount < 100 ~ "Less than $100",
        amount >= 100 & amount < 500 ~ "Between $100 and $499",
        amount >= 500 & amount < 1500 ~ "Between $500 and $1499",
        amount >= 1500 & amount < 3000 ~ "Between $500 and $2999",
        amount >= 3000 ~ "$3,000 or more"
      )
  )
```

I used this example to mutate my own data and create categories for undergraduate and graduate classes.

```{r}
umd_course_levels <- umd_separated |>
  mutate(
    course_code_category = case_when(
        course_code_num < 500 ~ "undergraduate",
        course_code_num >= 500 & course_code_num < 900 ~ "graduate")
  )
```

Here, I'm telling R to:

-   create a new data set from "umd_separated" — the new data set I just created to separate the course code numbers from the characters — and call the new data set "umd_course_levels"

-   prepare to mutate some data in a new column called "course_code_category"

-   in the new "course_code_category" column:

    -   categorize all the numbers below 500 in the "course_code_num" column as "undergraduate"

    -   categorize all the numbers between 500 and 900 in the "course_code_num" column as "graduate"

OK, and to actually answer the question, I used what we learned in class:

```{r}
umd_course_levels |>
  group_by(course_code_category) |>
  summarise(
    count_course_level = n(),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(median_seats))
```

```{r}
umd_course_levels |>
  group_by(course_code_category, department) |>
  summarise(
    count_courses = n(),
    mean_seats = mean(seats),
    median_seats = median(seats)
  ) |>
  arrange(desc(median_seats))
```

In the first chunk, I'm telling R to:

-   look at the data set "umd_course_levels"

-   group the data by the course code category

-   summarize the data by:

    -   counting the number of graduate and undergraduate classes

    -   calculating the mean and median number of seats for undergraduate and graduate classes

In the second chunk, everything is the same EXCEPT I'm telling R to group the data by both course_code_category and department.

In sum:

-   **How many of the classes are 500-level classes or higher?**

    -   24,007

-   **Do graduate classes tend to be smaller, larger or around the same size as undergraduate classes?**

    -   Graduate-level classes, which have an average of 18 seats and a median of 12 seats, tend to be much smaller than undergraduate-level classes, which have an average of 59 seats and a median of 30 seats.

-   **Does this differ by department?**

    -   Yes. While most departments offer larger undergraduate classes and smaller graduate classes, the graduate courses in some departments — like the Data Science & Analytics department — are larger than the undergraduate courses. Likewise, there are some departments — like the Greek and Music Performance departments — that have incredibly small undergraduate classes with a median of 5 seats.
