---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
readRenviron("~/.Renviron")
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2022 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2022 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
v21 <- load_variables(2021, "acs5", cache=TRUE)
```

## Answer questions

#### **Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first (search for it in the 'concept' column at the data frame we just created).

```{r}
md_median_age <- get_acs(geography = "county",
              variables = c(median_age = "B01002_001"),
              state = "MD",
              year = 2021)

md_median_age <- md_median_age |>
  rename(median_age=estimate) |>
  select(-variable)

md_median_age |>
  arrange(median_age)

```

**Answer here: Baltimore City has the lowest *estimated* median age at 35.7 years. (Unless you're asking a trick question, in which case I'll add that Wicomico County *COULD* have the lowest median age, because the margin of error would put the median age between 35.5 and 36.5. In other words, given the overlapping margins of error, there is a possibility that Wicomico County has a lower median age than Baltimore City.**

------------------------------------------------------------------------

#### **Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. **A2**

```{r}
md_median_age |>
  filter(median_age >= 50)

```

**Answer here: Talbot County (51 years) and Worcester County (50.4 years)**

------------------------------------------------------------------------

#### **Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1.  *Load the Montgomery County 911 overdose calls data.*
2.  *Create a new dataframe with the total number of calls for each zip code.*
3.  *Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).*
4.  *Get from the ACS data from all Maryland zip codes using that variable.*
5.  *Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.*
6.  *Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.*

Which zip code has the highest rate of calls? Which has the lowest?

```{r}
montgomery_od_911 <- read_csv("data/montgomery_2022_overdoses.csv")

montgomery_od_zip_overview <- montgomery_od_911 |>
  group_by(zip) |>
  summarise(calls=n()) |>
  arrange(desc(calls))

montgomery_od_zip_overview <- montgomery_od_zip_overview |>
  mutate(zip = as.character(zip))


md_population <- get_acs(geography = "zcta",
              variables = c(population_18_over = "B09021_001"),
              state = "MD",
              year = 2019) |>
  rename(
    population_18_over=estimate,
    zcta=NAME) |>
  select(-variable, -GEOID)
  
md_population <- md_population |>
  mutate(
    zcta = gsub("ZCTA5 ", "", zcta)
  )

montgomery_od_911_with_zcta <- montgomery_od_zip_overview |>
  left_join(md_population, join_by(zip==zcta))

montgomery_od_911_with_zcta <- montgomery_od_911_with_zcta |>
  mutate(calls_per_capita = calls/population_18_over * 1000)

montgomery_od_911_with_zcta |>
  arrange(desc(calls_per_capita))
```

**Answer here: ZIP code 20877 — the Gaithersburg area — had the highest per capita rate of 911 overdose calls at 1.7 calls per 1,000 people. ZIP 20707 — the Laurel area, where there was only one recorded 911 overdose call — had the lowest per capita call rate at 0.04 calls per 1,000 people.**

------------------------------------------------------------------------

#### **Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes on Q3 with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**Answer here:**

20877:

-   38,524 people (20% white, 18% Black, 16% Asian, 3% two or more races, **44% Hispanic**)

-   \$35,947 per capita income

-   13.2% poverty rate

-   **46.2% foreign-born**

20707:

-   36,352 people (20% white, 46% Black, 11% Asian, 6% two or more races, **16% Hispanic**)

-   \$49,071 per capita income

-   8.9% poverty rate

-   **28.3% foreign-born**

My answer is that there is *probably* a story here, but it would require more data. The possible story would likely center around the noticeable racial and citizenship disparities between the two zip codes. Roughly 44% of the residents in the Gaithersburg area are Hispanic — more than double the state average — compared to 16% in the Laurel area. Likewise, more than 46% of residents in the Gaitherburg area are immigrants, compared to 28% in the Laurel area. The reason I say that there is *probably* a story here is because there is no available language data. I know from my own NARCAN training that providing overdose prevention information and resources to Spanish-speaking communities can be difficult, and this tends to result in higher rates of overdoses in these communities. Given the noticeable racial and citizenship disparities between the zip codes, I would be curious to learn more about the potential language barriers this may present. I would also want to see fatal overdose demographic data from each of these two areas. All that being said, I'm relatively confident there is a story here.

------------------------------------------------------------------------

#### **Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

```{r}
montgomery_od_911_with_zcta <- montgomery_od_911_with_zcta |>
  mutate(
    min_pop_18_over = population_18_over-moe,
    max_pop_18_over = population_18_over+moe,
    min_pop_calls_per_capita = calls/min_pop_18_over * 1000,
    max_pop_calls_per_capita = calls/max_pop_18_over * 1000,
    )

montgomery_od_911_with_zcta
```

**Answer here: Having this context makes me more confident in my answers to Q3 because the margins of error don't even come close to overlapping for either ZIP. That is, ZIP 20707's highest possible per capita call rate is still significantly lower than the next-lowest ZIP code's lowest possible rate; the same is true for ZIP 20877, where the lowest possible per capital call rates is signifcantly higher than the next-highest ZIP code's highest possible rate.**
