---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
options(scipen=999)
# Load libraries
library(tidyverse)
library(janitor)
library(tidycensus)
library(tigris)
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history in 2022 by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1. The story here is that while Moore outperformed Brown — usually by a significant margin — in the vast majority of Maryland counties, there are two notable exceptions: PG County and Baltimore City. In each of these counties, Moore received a lower share of the vote than Brown. This is interesting because it begs the question why. Why were voters in only PG County and Baltimore City less inclined to vote for Moore than other Maryland voters? This would be the story to continue investigating.**

```{r}
#Calculate percentages for both candidates and the difference between them per county

md_electoral <- read_csv("data/md_statewide_county.csv") |>
  clean_names() |>
  mutate(geoid = as.character(geoid))

md_electoral_with_pct <- md_electoral |>
  group_by(county) |>
  mutate(
    gov_total = sum(cox, moore, lashar, wallace, harding, sefcik, gov_writein),
    ag_total = sum(peroutka, brown, ag_writein),
    pct_moore = moore/gov_total*100,
    pct_brown = brown/ag_total*100,
    pct_diff = pct_moore - pct_brown
  )
```

```{r}
#Take county geometry from tigris
counties <- counties()

#Filter out MD counties
md_counties <- counties |>
  filter(STATEFP == "24") |>
  clean_names()
```

```{r}
#Merge geometries with voting data
md_electoral_with_geometry <- md_counties |>
  left_join(md_electoral_with_pct, join_by(geoid))
```

```{r}
#Create the map
ggplot() +
  geom_sf(data=md_electoral_with_geometry, aes(fill=pct_diff)) +
  scale_colour_viridis_b(option="magma") +
  theme_minimal()
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2. The obvious question to ask here is, what happened in Talbot County in March 2023? The outage rate there is not only higher but *significantly* higher than the rate in any other county. Of course, reporting on this would require more data — for all we know, March 2023 was just a fluke. It's entirely possible that a storm that hit Talbot County particularly hard, and the outage rate there in other months is comparable to that of other counties. However, assuming the data show this was not a fluke, reporting this out would require finding out why and how long this has been a problem in Talbot County. For example, does the data show that this only became a problem during COVID? Does the onset of outage issues correspond with any major construction projects, storms, issues with the provider, etc.? All of these things could help get to the bottom of *why* this is a problem in Talbot County.**

```{r}
#Read data and calculate rate of outages per customer
county_outages <- read_csv("data/county_outages_march2023.csv") |>
  clean_names() |>
  mutate(outage_rate = outage_customers/total_customers*10000)
```

```{r}
#Merge geometries with outage data
md_counties <- md_counties |>
  mutate(
    namelsad = str_to_title(namelsad),
    namelsad = gsub(" County", "", namelsad)
  )

county_outages_with_geometry <- md_counties |>
  full_join(county_outages, join_by("namelsad"=="county"))
```

```{r}
#Map it out
ggplot() +
  geom_sf(data=county_outages_with_geometry, aes(fill=outage_rate)) +
  scale_fill_viridis_b(option="magma",trans = "log") +
  theme_minimal()
```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change.What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3. The percentage change map shows that the number of EVs in Caroline County grew at a significantly faster rate between 2020 and 2023 than in other Maryland counties. That is, all counties saw a more than 100% increase in the number of EVs between 2020 and 2023, but Caroline County experienced a whopping 425% increase — more than 150 percentage points higher than the county with the next-highest rate of EV growth. I would be a little cautious about drawing conclusions solely from this map, though, because Caroline County had the third-fewest EVs in 2020. In other words, a 400% increase in EVs isn't as hard to achieve when you only started with 27. Likewise, the per capita map clearly shows that the per capita rate of EVs in Caroline County is on the lower end compared to most other Maryland counties. This second map shows that Montgomery, Howard, Anne Arundel and Frederick counties have highest per capita rates of EVs. This makes sense — these are among the richest counties in the state. The second map also makes clear that the first map is a bit misleading, as the counties with the highest per capita rates of EVs still experienced growth rates north of 130%. I guess what I'm trying to say is that the first map highlights the EV growth rate in Caroline County but doesn't contextualize it, and the second map gives a much wider view of the state's EV picture as a whole.**

```{r}
#Read data and calculate change of EV vehicles
ev_data <- read_csv("data/ev_2020_2023.csv") |>
  mutate(pct_change = ((march_2023-july_2020)/july_2020)*100)
```

```{r}
#Merge geometries with EV data
ev_with_geometry <- md_counties |>
  full_join(ev_data, join_by("namelsad"=="county"))
```

```{r}
#Map out the change
ggplot() +
  geom_sf(data=ev_with_geometry, aes(fill=pct_change)) +
  scale_colour_viridis_b(option="magma") +
  theme_minimal()
```

```{r}
#Grab population data from Census
v21 <- load_variables(2021, "acs5", cache=TRUE)

md_population <- get_acs(geography = "county",
              variables = c(total_population = "B01003_001"),
              state = "MD",
              year = 2021)

ev_with_geometry_and_population <- ev_with_geometry |>
  full_join(md_population, join_by("geoid"=="GEOID"))
```

```{r}
#Calculate per capita rate for March 2023
ev_with_geometry_and_population <- ev_with_geometry_and_population |>
  mutate(ev_per_capita = march_2023/estimate*100)
```

```{r}
#Map out the per capita rate
ggplot() +
  geom_sf(data=ev_with_geometry_and_population, aes(fill=ev_per_capita)) +
  scale_colour_viridis_b(option="magma") +
  theme_minimal()
```
